Relationships in the extremes and their influence on competition and coexistence

### 2021-09-13 paper/Paper.Rnw
starting edits to Intro

%***Add to these
\keywords{kw1, kw2,kw3}

@@

%Old version, prior to my edits: Central to ecology is the maintenance of biodiversity in competitive ecosystems. Populations subject to strong
%competitive pressure may succumb to competitive exclusion - an idea which postulates if species within the same
%trophic level and similar niches exist, one will eventually outcompete the other (Hardin, 1960). Mediation of
%competitive pressure and exclusion allow for species to stably coexist. The criteria for coexistence and competitive
%exclusion can be broken down into contributing mechanisms provided by modern coexistence theory, a framework
%which seeks to assess and quantify mechanisms that result in coexisting species and thus the maintenance of
%biodiversity (Chesson, 1994).
%Now the new, edited version, with some additional changes requested, to be made by Jasmin, please:
The maintenance of biodiversity in competitive ecosystems is a research topic at the absolute core of modern ecology \citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
%***DAN: Jasmin, add a few classic references. See the first sentence of Ellner et al 2019, read the abstracts of those papers
%(more than the abstracts if you want), and consider using them as references here.
The competitive exclusion principle, covered, for instance, in most or all introductory ecology courses and texts, states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will outcompete 
the other and drive it extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how stable species coexistence 
nevertheless occurs, and biodiversity in maintained, in real ecosystems, through stabilizing mechanisms and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with name such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to quite a few empirical systems, including annual plants \citep{Angert_2009, Hallett_2019}, perennial plants \citep{Adler_2010, Adler_2006}, tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}, 
%***DAN: 
%Please also read the abstract of the Angert reference to make sure it is appropriate to cite it in this way (and so that
%you become aware of that reference). One place to look for more references to put in for Y and Z is the top of the 
%right column on the first page of the Barabas review (Ecol Monogr, 2018).
even though it is mathematically complex in the eyes
of most ecologists (but see \cite{Ellner_2016,Ellner_2019} for efforts to make it more readily applied).
%***DAN: Jasmin, note the differences carefully, please, between your version (commented, above) and the edits I've 
%made. It's the same in main structure, making all the same points. It's just a bit more fleshed out. One of the purposes
%of an Intro is to link the reader to the appropriate background lit. You need a fair number of references and a little
%more detail than you previously gave to do that.

%Old version, some comments interspersed: 
%Equally central to ecology is the study of relationships between ecological variables; however, it is almost exclusively blanketed by simple measures of associativity, i.e., correlation coefficients.
%***DAN: "Blanketed" is a weird verb in this context. 
%***DAN: Also, MCT is a conceptually deep theory, and ATA is a fairly simple observation that happens to have some consequences. I would not be confortable with phrasing that seems to put them on the same footing! 
%Although two sets of variables may have equal correlation coefficients, they can show different association structures (see Figure).
%More specifically, the association within the tail ends of data and the asymmetry between the tail associations, termed asymmetric tail associations (ATAs), tell a more interesting story about the relationship between variables...and?...
%Recent work has shown that ATAs have important considerations for ecological phenomena; although, the implications of ATAs for competition and coexistence have not yet been investigated. 
%Asymmetry of tail associations within a dataset of variables describes uneven correlations between the tails of a dataset.
%For a simple example, a dataset with more association in the lower (upper) half and lesser association in the upper (lower) half is termed left (right) tail associated. 
%The application of ATAs in ecology is relatively new, and in exploration of ATAs in ecology, Gosh et al (AER) found ATAs in all eight ecological datasets they considered.
%Thus, the existence and functioning role of ATAs in nature is reasonable to assume and question. 
%In fact, ATAs can have influence on ecosystem service stability, Taylor’s law, and extinction risks via extreme climatic events. Conversely, ATA’s can be influenced by Moran Effect and Leibeig’s Law of Minimum \citep{Ghosh_2020_Eco, Ghosh_2020_AER}.
%Despite the frequent occurrence of ATAs in ecological data, the implications of tail associations for competition and coexistence, to our knowledge, has not yet been examined; doing so is the main goal of this paper.
%New version:
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between variables provide only a small portion of the overall information content contained in the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in figure \ref{fig:noise} all have the same
Pearson correlation. 
But each panel of figure \ref{fig:noise} nevertheless
depicts a very different type of association. Other metrics of association are used, in addition to correlations
(e.g., various common forms of regression), but correlations are extremely common, and other common
methods also provide only a small portion of the available information \citep{Nelsen2006}.
The variables in figure \ref{fig:noise}A,C 
%***
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (figure \ref{fig:noise}A, termed ``left-tail association'') or right (figure \ref{fig:noise}C,
termed `right-tail association') portions of the variables' distributions. Although ATAs and related ideas had previously been very seldom 
considered in ecology and related areas (but see \citep{Anderson_2019, deValpine_2014, Popovic_2019}), 
%***
%
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can be mechanistically important for 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.
%***DAN: Again note the similarity between the old and new versions, the new is just a bit more fleshed out, and maybe a
%bit more precise in some of the wordings in places.

%***DAN: Please combine the following two paragraphs into one single paragraph with this topic sentence:
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence, in particular: storage effects. We review, conceptually, what storage effects are and we explain 
why ATAs may be relevant to them.
%***DAN: I have inserted some comments below that might help.
Storage effects allow for species to fluctuate over time while escaping extinction, continously recovering from low densities. 
In order for storage effects to contribute as a stabilizing mechanism to coexistence, there must first be covariation between competitive pressures and environmental fluctuations [EC covariance \citep{Chesson_1994, Barabas_2018}].
%Because the contributive value of storage effects relies on the relationship between environment and competition \citep{Chesson_1994, Barabas_2018}, it reasonably follows to suspect the contribution of storage effectsto be altered by an association between environment and competition that is asymmetric, i.e., ATAs between environment and competition.
The reliance of storage effect's magnitude on the degree of relationship between environment and competition indicate that storage effects involve environment-coexistence associations, and hence there is potential for ATAs to influence storage effects. 
In addition to EC covariance, functioning temporal storage effects also require that heterospecific populations respond differently to environmental fluctuations, and different life stages must also vary in responses \citep{}. 
In a simple two species system, differences in environmental response allow for one species to grow to high densities in its favorable environemnt while the other is very low. %**JAS: should we introduce invader analysis or terminology here?
Growth is stable for the high density species due to positive EC covariance -- high intraspecific competition limits its growth rate in a favorable environment. 
Conversely, EC covariance is lesser for the low density species, leading to large fluctuations in population growth rate in both directions. 
Although, this may appear like a scenario for extiction, it can, in fact, be advantageous due to differences between life stages, buffering species against extinction by limiting negative growth rates, but not positive ones. 
For example, insensitive adults or dormants seeds can still persist when in an unfavorable environment and overall growth rate is negative \citep{Chesson_2012}; environment and competiton do not effect the buffered life stages.   
Thus, averaged over multiple generations, higher fluctuation in growth rate can lead to a positive low density growth rate and coexistence is possible \citep{Chesson_2000}.
Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events.

@@

%move this paragraph to discussion
%Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
%However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
%All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
%Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
%***DAN: I added the above sentence. Not a fully fledged "goal" but a serious thing
%to discuss.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.

>>>>

%***Add to these
\keywords{competition, coexistence, biodiversity, modern theory of coexistence, storage effects, asymmetric tail association}

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity in maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to multiple systems, including annual plants \citep{Angert_2009, Hallett_2019}, perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although the MTC is mathematically complex to most ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in figure \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also used, in addition to correlations
(e.g., various common forms of regression), but correlations are extremely common, and other common
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in figure \ref{fig:noise}A,C 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (figure \ref{fig:noise}A, termed ``left-tail association'') or right (figure \ref{fig:noise}C,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs and related ideas had previously been very seldom 
considered in ecology and related areas (but see \citep{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence, in particular: storage effects. We review, conceptually, what storage effects are and we explain 
why ATAs may be relevant to them.
%***
Storage effects allow for species to fluctuate over time while escaping extinction, continously recovering from low densities. 
In order for storage effects to contribute as a stabilizing mechanism to coexistence, there must first be covariation between competitive pressures and environmental fluctuations [EC covariance \citep{Chesson_1994, Barabas_2018}].
%Because the contributive value of storage effects relies on the relationship between environment and competition \citep{Chesson_1994, Barabas_2018}, it reasonably follows to suspect the contribution of storage effectsto be altered by an association between environment and competition that is asymmetric, i.e., ATAs between environment and competition.
The reliance of storage effect's magnitude on the degree of relationship between environment and competition indicate that storage effects involve environment-coexistence associations, and hence there is potential for ATAs to influence storage effects. 
In addition to EC covariance, functioning temporal storage effects also require that heterospecific populations respond differently to environmental fluctuations, and different life stages must also vary in responses \citep{}. 
In a simple two species system, differences in environmental response allow for one species to grow to high densities in its favorable environemnt while the other is very low. %**JAS: should we introduce invader analysis or terminology here?
Growth is stable for the high density species due to positive EC covariance -- high intraspecific competition limits its growth rate in a favorable environment. 
Conversely, EC covariance is lesser for the low density species, leading to large fluctuations in population growth rate in both directions. 
Although, this may appear like a scenario for extiction, it can, in fact, be advantageous due to differences between life stages, buffering species against extinction by limiting negative growth rates, but not positive ones. 
For example, insensitive adults or dormants seeds can still persist when in an unfavorable environment and overall growth rate is negative \citep{Chesson_2012}; environment and competiton do not effect the buffered life stages.   
Thus, averaged over multiple generations, higher fluctuation in growth rate can lead to a positive low density growth rate and coexistence is possible \citep{Chesson_2000}.
Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events.

@@

%move this paragraph to discussion
%Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
%However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
%All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
%Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.
